---
title: "Week 10 Assignment"
output: html_document
author: 'Will Berritt'
date: "2024-04-01"
---

# OBJECTIVE: Perform some sort of sentiment analysis on a group of text. I found a list of titles to run analysis on. Curious to see if I can associate a sentiment score to the titling of a specific author.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Load packages and create unnested data set
```{r, message=FALSE}
# Load packages
library(gutenbergr)
library(dplyr)
library(tidytext)

# Load Gutenberg metadata
gute <- gutenberg_metadata

# Tokenize titles
gute_unnest <- gute |> unnest_tokens(word, title)
head(gute_unnest, 20)
```
### Remove stop words and join with 'bing' sentiment lexicon
```{r, message=FALSE}
# Remove stop words
gute_clean <- gute_unnest |> anti_join(stop_words)
head(gute_clean, 20)

# Join with sentiment lexicon
sentiment_join <- gute_clean |> inner_join(get_sentiments('bing'))
head(sentiment_join, 20)
```
### See how many distinct sentiments there are
```{r, message=FALSE}
# Count distinct sentiments
n_distinct(sentiment_join$sentiment)
```
### Assign numerical scores to positive and negative sentiments
```{r, message=FALSE}
# Assign sentiment scores
sentiment_w_score <- sentiment_join %>% 
  mutate(sentiment_score = if_else(sentiment == 'positive', 1, -1))
head(sentiment_w_score, 20)
```
### Remove nulls and create mean sentiment scores by author
```{r, message=FALSE}
# Remove NAs
sentiment_w_score <- na.omit(sentiment_w_score)

# Calculate average sentiment score by author
mean_score_by_author <- sentiment_w_score |> group_by(author) |> 
  summarize(avg_sentiment_score = mean(sentiment_score), count = n())
head(mean_score_by_author, 20)
```
### Filter out to only authors with 50 or more titles
```{r, message=FALSE}
# Filter authors with at least 50 titles
most_popular <- subset(mean_score_by_author, count >= 50)

# Arrange by average sentiment score
most_popular |> arrange(avg_sentiment_score)
head(most_popular, 20)
```
### Instead of running sentiment analysis on total amount of text in the titles from different authors because it will skew to the longer titles - I want to group by title then by author so each title is weighed equally
```{r, message=FALSE}
# Calculate average sentiment score by Gutenberg ID
new_sentiment_join <- sentiment_w_score |> group_by(gutenberg_id) |> 
  summarize(avg_sentiment_score = mean(sentiment_score), count = n())
head(new_sentiment_join, 20)
```
### Subset to verify there are scores between [-1,1]
```{r, message=FALSE}
# QA: Filter sentiment scores within range
subset(new_sentiment_join, avg_sentiment_score > -1 & avg_sentiment_score < 1 & avg_sentiment_score != 0)
```
### Join data frames and omit nulls
```{r, message=FALSE}
# Left join with Gutenberg metadata
new_df <- left_join(new_sentiment_join, gute, by = 'gutenberg_id')
new_df$author <- na.omit(new_df$author)
head(new_df, 20)
```
### Create mean score by author based on title sentiment score
```{r, message=FALSE}
# Calculate average title sentiment score by author
average_title_by_author <- new_df |> group_by(author) |> 
  summarize(title_sentiment_score = mean(avg_sentiment_score), count = n())
head(average_title_by_author, 20)
```
### Filter only authors with 25+ titles
```{r, message=FALSE}
# Filter authors with at least 25 titles
author_sentiment <- subset(average_title_by_author, count >= 25)

# Arrange by title sentiment score
author_sentiment |> arrange(title_sentiment_score)
head(author_sentiment, 20)
```
### QA why this author is perfectly -1
```{r, message=FALSE}
# Subset Gutenberg metadata by specific author
subset(gute, author == 'United States. Work Projects Administration')
```
### View most positive authors
```{r, message=FALSE}
# Arrange author sentiment in descending order
author_sentiment |> arrange(desc(title_sentiment_score))
head(author_sentiment, 20)
```
### QA why Weymouth is excatly 1
```{r, message=FALSE}
# Subset Gutenberg metadata by specific author
subset(gute, author == 'Weymouth, Richard Francis')
```

# Conclusion: Interesting to see trends amongst authors. Generalizing a sentiment score to an author like Mark Twain is pretty cool and alludes to the power of programmatic sentiment analysis. An interesting project I thought of using a similar skill set would be to take media outlets and assign sentiment scores based on the article titles they publish - i.e. NY Times has an average score of -0.5, while The Economist has a sentiment score of 0.2 (these are not real values but just examples) and categorize as positive or 'happy' media outlets vs. negative or 'depressing' media outlets.

### References: 
### Robinson, David. 2016. gutenbergr: Download and Process Public Domain Works from Project Gutenberg. https://cran.rstudio.com/package=gutenbergr.
