---
title: "will_analysis_proj_3"
output: html_document
date: "2024-03-18"
---
# Intro: After first connecting to the normalized Azure database and creating a main data frame to work with, I wanted to perform analysis and assess how each independent variable affects the pay one can expect. What's the most important factors? What differences can help us understand which roles will pay more than others? How do different experience levels affect income? Job titles? How have salaries changed from 2020 to 2022? Has it outpaced inflation? With these questions in mind I began my analysis. 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Load packages
```{r, message = FALSE}
# Load libraries
library('DBI')  # Database interface in R
library('RMySQL')  # MySQL driver for R
library('tidyr')  # Data tidying functions
library('dplyr')  # Data tidying functions
library(ggplot2)
```
### Create dataframe for analysis from Azure
```{r}
# Connect to SQL DB in R
mydb <- dbConnect(MySQL(), user='chhiring.lama65', password='lama65', dbname='chhiring.lama65', host='cunydata607sql.mysql.database.azure.com')

# Retrieve data from database tables
job <- dbGetQuery(mydb,'select * from job')
company <- dbGetQuery(mydb,'select * from company')

# Join tables based on a common column 'cid'
total_df_joined <- left_join(company, job, by='cid')

# Join tables and filter data
total_df <- left_join(company, job, by='cid')
total_df <- subset(total_df, total_df$employment_type=='FT')

# Count null rows in the dataframe
count_null_rows <- sum(rowSums(is.na(total_df)) > 0)
```
### Analyze pay by company
```{r, message=FALSE}
# Calculate mean and median salary by company
mean_salary_by_company <- total_df |> group_by(cid, salary_currency) |>  # Group by 'cid' and 'salary_currency'
  summarize(mean_salary = mean(salary_in_usd), median_salary = median(salary_in_usd), count = n())  # Calculate mean and median salaries, and count of records
result_by_company <- mean_salary_by_company |> arrange(desc(mean_salary))  # Arrange by descending mean salary
result_by_company <- subset(result_by_company, result_by_company$count>5)  # Subset companies with more than 5 records

# Add cost of living information
eur_avg_monthly_cost <- mean(c(1260, 1128, 1414))  # Calculate average monthly cost for EUR currency
cost_of_living <- tibble(
  salary_currency = c("USD", "GBP", "EUR", "INR"),  # Currency codes
  monthly_cost = c(1951, 1929, eur_avg_monthly_cost, 423)  # Monthly costs for each currency
)
w_monthly_cost <- left_join(result_by_company, cost_of_living, by='salary_currency')  # Join salary and cost of living data
w_monthly_cost$annual_cost <- w_monthly_cost$monthly_cost*12  # Calculate annual cost
w_monthly_cost$proportion_income_monthly_exp <- w_monthly_cost$annual_cost/w_monthly_cost$mean_salary  # Calculate wealthiness based on annual cost and mean salary

# Arrange companies by wealthiness
wealthiness <- w_monthly_cost |> arrange(proportion_income_monthly_exp)  # Arrange by descending wealthiness
wealthiness[c("cid","salary_currency", "mean_salary","proportion_income_monthly_exp")]
```
### Analyze pay by job title
```{r, message=FALSE}
# Calculate mean and median salary by job title
mean_salary_by_job_title <- total_df |> group_by(job_title) |> 
  summarize(mean_salary = mean(salary_in_usd), median_salary = median(salary_in_usd), count = n())

# Arrange job titles by descending mean salary
result_job_title <- mean_salary_by_job_title |> arrange(desc(mean_salary))
result_job_title <- subset(result_job_title, result_job_title$count>5)
result_job_title

# Perform sub-analysis for specific job titles and locations
engineer_consultant_country <- subset(total_df, total_df$job_title=='Big Data Engineer' | total_df$job_title=='Data Science Consultant') |> 
  group_by(company_location) |> summarize(mean_by_country_title = mean(salary_in_usd), median = median(salary_in_usd), count = n()) |> arrange(desc(mean_by_country_title))
engineer_consultant_country
```
### Analyze pay by remote ratio
```{r, message=FALSE}
# Calculate mean salary by remote work ratio
mean_salary_by_remote_ratio <- total_df |> group_by(remote_ratio) |> 
  summarize(mean_salary = mean(salary_in_usd), median_salary = median(salary_in_usd), count = n())

# Arrange remote work ratios by descending mean salary
result_remote_ratio <- mean_salary_by_remote_ratio |> arrange(desc(mean_salary))
result_remote_ratio <- subset(result_remote_ratio, result_remote_ratio$count>5)
result_remote_ratio

# Perform sub-analysis for specific remote work ratios and company locations
mean_salary_by_remote_ratio <- total_df |> group_by(remote_ratio, company_location) |> 
  summarize(mean_salary = mean(salary_in_usd), median_salary = median(salary_in_usd), count = n())

# Arrange remote work ratios by descending mean salary
result_remote_ratio <- mean_salary_by_remote_ratio |> arrange(desc(mean_salary))

# Subset remote work ratios with specific values and counts
result_remote_ratio_100 <- subset(result_remote_ratio, remote_ratio==100 & count>2)
result_remote_ratio_50 <- subset(result_remote_ratio, remote_ratio==50 & count>2)
result_remote_ratio_0 <- subset(result_remote_ratio, remote_ratio==0 & count>2)
result_remote_ratio_100
result_remote_ratio_50
result_remote_ratio_0
```
### Analyze pay by currency offered
```{r}
# Calculate mean salary by offered currency
mean_salary_by_currency_offered <- total_df |> group_by(salary_currency) |> 
  summarize(mean_salary = mean(salary_in_usd), median_salary = median(salary_in_usd), count = n())

# Arrange currencies by descending mean salary
result_currency_offered <- mean_salary_by_currency_offered |> arrange(desc(mean_salary))
result_currency_offered <- subset(result_currency_offered, result_currency_offered$count>5)
result_currency_offered
```

### Analyze pay by year
```{r}
# Calculate mean salary by year
mean_salary_by_year <- total_df |> group_by(work_year) |> 
  summarize(mean_salary = mean(salary_in_usd), median_salary = median(salary_in_usd), count = n())

# Arrange years by ascending mean salary
result_by_year <- mean_salary_by_year |> arrange(mean_salary)
result_by_year <- subset(result_by_year, result_by_year$count>5)
result_by_year

# Add inflation data
inflation <- tibble(
  work_year = c(2020, 2021, 2022),
  inflation_rate = c(1.40, 7.00, 6.50)
)
results_inflation <- left_join(result_by_year, inflation, by = 'work_year')

# Figure out salary growth vs inflation
results_inflation <- results_inflation |> mutate(percent_change = (mean_salary / lag(mean_salary) - 1) * 100)
perfomance_vs_inflation <- (125195)/((1 + 1.40 / 100)*(1 + 7 / 100)*(1 + 6.5 / 100)*(98543))
perfomance_vs_inflation <- (perfomance_vs_inflation - 1)*100
paste('Salary growth has outperformed inflation rate by ', perfomance_vs_inflation, '% since 2020', sep = '')
```

### Analyze pay by experience level
```{r}
# Calculate mean salary by experience level
mean_salary_by_exp_level <- total_df |> group_by(experience_level) |> 
  summarize(mean_salary = mean(salary_in_usd), median_salary = median(salary_in_usd), count = n())

# Arrange experience levels by descending mean salary
result_exp_level <- mean_salary_by_exp_level |> arrange(desc(mean_salary))
result_exp_level <- subset(result_exp_level, result_exp_level$count>5)
result_exp_level

# Define years of experience for each level
yoe <- tibble(
  experience_level = c('EX', 'SE', 'MI', 'EN'),
  year_of_experience = c(10, 8, 4, 1)
)

# Join experience levels with years of experience
exp_leve_w_years <- left_join(result_exp_level, yoe, by = 'experience_level')
exp_leve_w_years <- exp_leve_w_years |> arrange(mean_salary)
exp_leve_w_years <- exp_leve_w_years |> arrange(mean_salary)
exp_leve_w_years

# Calculate mean salary changes and average pay increase by year
exp_leve_w_years <- exp_leve_w_years |> 
  mutate(mean_salary_change = mean_salary - lag(mean_salary, default = first(mean_salary)),
         year_of_experience_change = year_of_experience - lag(year_of_experience, default = first(year_of_experience)))
exp_leve_w_years$mean_salary_change_average <- exp_leve_w_years$mean_salary_change/exp_leve_w_years$year_of_experience_change
average_pay_increase_by_year <- mean(exp_leve_w_years$mean_salary_change_average, na.rm = TRUE)
average_pay_increase_by_year

# Reshape data into long format
data_long <- pivot_longer(exp_leve_w_years, cols = c(mean_salary, median_salary),
                          names_to = "salary_type", values_to = "salary")

# Create a grouped bar chart
ggplot(data = data_long, aes(x = experience_level, y = salary, fill = salary_type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), color = "black") +
  labs(title = "Mean and Median Salary by Experience Level",
       x = "Experience Level",
       y = "Income",
       fill = "Salary Type") +
  scale_fill_manual(values = c("mean_salary" = "blue", "median_salary" = "red")) +
  theme_minimal()
```
### Analyze pay by employee residence
```{r}
# Calculate mean salary by employee residence
mean_salary_by_employee_res <- total_df |> group_by(employee_residence) |> 
  summarize(mean_salary = mean(salary_in_usd), median_salary = median(salary_in_usd), count = n())
result_employee_res <- mean_salary_by_employee_res |> arrange(desc(mean_salary))
result_employee_res <- subset(result_employee_res, result_employee_res$count>5)
result_employee_res
```
### Analyze pay by company location
```{r}
# Calculate mean salary by company location
mean_salary_by_company_loc <- total_df |> group_by(company_location) |> 
  summarize(mean_salary = mean(salary_in_usd), median_salary = median(salary_in_usd), count = n())
result_company_loc <- mean_salary_by_company_loc |> arrange(desc(mean_salary))
result_company_loc <- subset(result_company_loc, result_company_loc$count>5)
result_company_loc
```

# Conclusion: There are many different insights this analysis provides but there are a few extra-interesting points of analysis: 
### 1. The USA has the highest nominal and relative pay proportionate to monthly expenses. India, despite having the lowest mean_salary, has the next highest pay proportionate to monthly expenses - greater than Greece, Great Britain and Spain. 
### 2. 50% remote pays significantly less than both 0% remote and 100% remote
### 3. Salary growth in the data science industry has outperformed inflation by about 10% - which is good given high inflation rates. Granted, these are salaries for a variety of countries and continents so comparing directly to US inflation rates is not a perfect science
### 4. (After assigning a somewhat arbitrary but defensible number of years of experience to experience_levels) the yearly increase in pay spanning from entry level to executive level is between $15k-16k. This number is probably better represented by a perecentage but interesting to see what a year of experience is worth in the Data Science world.