---
title: "Project 2: Migration"
output: html_document
author: 'Will Berritt'
date: "2024-03-04"
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# # Introduction: This data set includes migration data data broken out by year and country. I aimed to analyze a few different aspects of this data:
1. Which countries have had the largest inflow of people? Which countries have had the most people leave?
2. Is there a realtionship between poverty and migration rates?
3. If we control for population growth, what do the relative migration rates look like?
4. What countries are leading in absolute migration? Broken out by decade?
With these questions in mind, I began working with and manipulating the data to extract these insights.

#### Load Packages
```{r, message=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(ggrepel)
```

#### Migration: Read Data Set number 1 and pivot the data
```{r, message=FALSE}
# Read CSV file, skip 3 rows, remove 'X' column
df1 <- read.csv("/Users/williamberritt/Downloads/API_SM/API_SM.POP.NETM_DS2_en_csv_v2_105.csv", header = TRUE, skip = 3) |> subset(select = -c(X))

# Reshape data from wide to long format using pivot_longer
df1 <- df1 |> pivot_longer(cols = (X1960:X2022), names_to = 'year', values_to ='value')

# Display first few rows of the reshaped data frame
head(df1)
```

#### Left Join with list of countries to remove all none country entities like continents, middle class, etc.
```{r, message=FALSE}
# Calculate total migration by country code
migration_by_country_code <- df1 |> 
  group_by(Country.Name) |> 
  summarize(total_migration = sum(value))

# Rename columns
colnames(migration_by_country_code) <- c("country", "total_migration")

# List of country names
country_names <- c("Afghanistan", "Albania", "Algeria", "Andorra", "Angola", "Antigua and Barbuda", "Argentina", "Armenia", "Australia", "Austria", "Azerbaijan", "Bahamas", "Bahrain", "Bangladesh", "Barbados", "Belarus", "Belgium", "Belize", "Benin", "Bhutan", "Bolivia", "Bosnia and Herzegovina", "Botswana", "Brazil", "Brunei", "Bulgaria", "Burkina Faso", "Burundi", "Cabo Verde", "Cambodia", "Cameroon", "Canada", "Central African Republic", "Chad", "Chile", "China", "Colombia", "Comoros", "Congo", "Costa Rica", "Croatia", "Cuba", "Cyprus", "Czech Republic", "Denmark", "Djibouti", "Dominica", "Dominican Republic", "Ecuador", "Egypt", "El Salvador", "Equatorial Guinea", "Eritrea", "Estonia", "Eswatini", "Ethiopia", "Fiji", "Finland", "France", "Gabon", "Gambia", "Georgia", "Germany", "Ghana", "Greece", "Grenada", "Guatemala", "Guinea", "Guinea-Bissau", "Guyana", "Haiti", "Honduras", "Hungary", "Iceland", "India", "Indonesia", "Iran", "Iraq", "Ireland", "Israel", "Italy", "Jamaica", "Japan", "Jordan", "Kazakhstan", "Kenya", "Kiribati", "Korea, North", "Korea, South", "Kosovo", "Kuwait", "Kyrgyzstan", "Laos", "Latvia", "Lebanon", "Lesotho", "Liberia", "Libya", "Liechtenstein", "Lithuania", "Luxembourg", "Madagascar", "Malawi", "Malaysia", "Maldives", "Mali", "Malta", "Marshall Islands", "Mauritania", "Mauritius", "Mexico", "Micronesia", "Moldova", "Monaco", "Mongolia", "Montenegro", "Morocco", "Mozambique", "Myanmar", "Namibia", "Nauru", "Nepal", "Netherlands", "New Zealand", "Nicaragua", "Niger", "Nigeria", "North Macedonia", "Norway", "Oman", "Pakistan", "Palau", "Panama", "Papua New Guinea", "Paraguay", "Peru", "Philippines", "Poland", "Portugal", "Qatar", "Romania", "Russia", "Rwanda", "Saint Kitts and Nevis", "Saint Lucia", "Saint Vincent and the Grenadines", "Samoa", "San Marino", "Sao Tome and Principe", "Saudi Arabia", "Senegal", "Serbia", "Seychelles", "Sierra Leone", "Singapore", "Slovakia", "Slovenia", "Solomon Islands", "Somalia", "South Africa", "South Sudan", "Spain", "Sri Lanka", "Sudan", "Suriname", "Sweden", "Switzerland", "Syria", "Taiwan", "Tajikistan", "Tanzania", "Thailand", "Timor-Leste", "Togo", "Tonga", "Trinidad and Tobago", "Tunisia", "Turkey", "Turkmenistan", "Tuvalu", "Uganda", "Ukraine", "United Arab Emirates", "United Kingdom", "United States", "Uruguay", "Uzbekistan", "Vanuatu", "Vatican City", "Venezuela", "Vietnam", "Yemen", "Zambia", "Zimbabwe")

# Create a data frame with country names
countries_df <- data.frame(country = country_names)

# Merge country names with migration data and remove rows with missing migration data
countries_only <- na.omit(left_join(countries_df, migration_by_country_code, by='country'))
```

#### Create data frames that have the countries with most amount of people coming or going
```{r, message=FALSE}
# Extract the top 5 countries with the highest migration influx
top_5_highest_influx <- head(countries_only[order(-countries_only$total_migration),])
top_5_highest_influx

# Extract the top 5 countries with the highest migration exodus
top_5_highest_exodus <- head(countries_only[order(countries_only$total_migration),])
top_5_highest_exodus
```

#### Read poverty data set
```{r, message=FALSE}
# Read Poverty Data Set
poverty <- read.csv("/Users/williamberritt/Downloads/poverty-rate-by-country-2024.csv")
head(poverty)
```

#### Merge poverty data with migration numbers to see relationship between migration and poverty
```{r, message=FALSE}
# Merge country names with poverty data
countries_poverty <- left_join(countries_df, poverty, by='country')

# Select relevant columns and remove rows with missing data
countries_poverty <- subset(countries_poverty, select = c('country', 'PovertyRateValue'))
countries_poverty <- na.omit(countries_poverty)

# Merge poverty data with migration data
migration_vs_poverty <- left_join(countries_poverty, migration_by_country_code, by='country')

# Remove rows with missing data
migration_vs_poverty <- na.omit(migration_vs_poverty)

# Extract variables for the plot
x <- migration_vs_poverty$PovertyRateValue
y <- migration_vs_poverty$total_migration

# Create a scatter plot
plot(x, y)
```

#### Plot cumulative migration by decade
```{r, message=FALSE}
# Create a copy of the original data frame
df_decade <- df1

# Add a column for the absolute value of migration
df_decade$absolute_value <- abs(df_decade$value)

# Extract the year from the 'year' column
df_decade$year <- str_replace(df_decade$year, 'X', '')
df_decade$year <- as.numeric(df_decade$year)

# Create a new column for the decade
df_decade$decade <- (floor(df_decade$year/10))*10

# Rename columns
colnames(df_decade) <- c("country", "code", 'indicate', 'indic_code', 'year', 'value', 'absolute_value', 'decade')

# Merge country names with the updated data frame
df_decade <- left_join(countries_df, df_decade, by='country')
df_decade <- na.omit(df_decade)

# Calculate total absolute migration by decade
migration_by_decade <- df_decade |> group_by(decade) |> summarize(absolute_migration = sum(absolute_value))

# Adjust the value for the decade 2020
migration_by_decade$adj_value <- ifelse(migration_by_decade$decade == 2020, migration_by_decade$absolute_migration/.3, migration_by_decade$absolute_migration)

# Create a plot
plot(migration_by_decade$decade, migration_by_decade$adj_value)
```

#### Adjust migration for population growth and observe differences
```{r, message=FALSE}
# Read population data from CSV file
pop_data <- read.csv('/Users/williamberritt/Documents/pop_data_proj2.csv')

# Select relevant columns
pop_data <- subset(pop_data, select = c('Year', 'World.Population'))

# Create a new column for the decade
pop_data$decade <- (floor(pop_data$Year/10))*10

# Remove commas from population numbers and convert to numeric
pop_data$World.Population <- str_replace_all(pop_data$World.Population, ',', '')
pop_data$World.Population <- as.numeric(pop_data$World.Population)

# Calculate average population by decade
avg_pop_decade <- pop_data |> group_by(decade) |> summarize(avg_pop = mean(World.Population))

# Merge average population and migration data by decade
migration_with_pop <- left_join(avg_pop_decade, migration_by_decade, by='decade')

# Calculate adjusted absolute migration rate per 100 people
migration_with_pop$adj_absolute_migration_rate_per_100 <- (migration_with_pop$adj_value / migration_with_pop$avg_pop) * 100

# Create a plot
plot(migration_with_pop$decade, migration_with_pop$adj_absolute_migration_rate_per_100)
```

#### Group data by decade and see which countries had highest absolute movement
```{r, message=FALSE}
# Group data by country and decade, calculating absolute migration
country_grouped_by_decade <- df_decade |> group_by(country, decade) |> summarize(absolute_migration = sum(absolute_value))

# Merge with country names
ONLY_country_grouped_by_decade <- left_join(countries_df, country_grouped_by_decade, by='country')
ONLY_country_grouped_by_decade <- na.omit(ONLY_country_grouped_by_decade)

# Subset data for each decade
decade_analysis_1960 <- subset(ONLY_country_grouped_by_decade, decade == 1960)
decade_analysis_1970 <- subset(ONLY_country_grouped_by_decade, decade == 1970)
decade_analysis_1980 <- subset(ONLY_country_grouped_by_decade, decade == 1980)
decade_analysis_1990 <- subset(ONLY_country_grouped_by_decade, decade == 1990)
decade_analysis_2000 <- subset(ONLY_country_grouped_by_decade, decade == 2000)
decade_analysis_2010 <- subset(ONLY_country_grouped_by_decade, decade == 2010)
decade_analysis_2020 <- subset(ONLY_country_grouped_by_decade, decade == 2020)

# Extract top 5 countries with the highest absolute migration for each decade
decade_analysis_1960 <- head(decade_analysis_1960[order(-decade_analysis_1960$absolute_migration),])
decade_analysis_1970 <- head(decade_analysis_1970[order(-decade_analysis_1970$absolute_migration),])
decade_analysis_1980 <- head(decade_analysis_1980[order(-decade_analysis_1980$absolute_migration),])
decade_analysis_1990 <- head(decade_analysis_1990[order(-decade_analysis_1990$absolute_migration),])
decade_analysis_2000 <- head(decade_analysis_2000[order(-decade_analysis_2000$absolute_migration),])
decade_analysis_2010 <- head(decade_analysis_2010[order(-decade_analysis_2010$absolute_migration),])
decade_analysis_2020 <- head(decade_analysis_2020[order(-decade_analysis_2020$absolute_migration),])

# Display results for 2020 and 1960
decade_analysis_2020
decade_analysis_1960
```

# Conclusion:
1. United States has gained the most migrants by a wide margin since 1960. On the other hand, Bangladesh has seen the largest exodus since 1960, closely followed by Pakistan, China and Mexico.
2. When looking at poverty rates and migration, we don't see much of a pattern outside of the fact that the countries experiencing the higest level of migration are lower in poverty.
3. Controlling for population growth when assessing migration rate is critical! The numbers will show that migration has trended positive in the last 60 years but if you factor in population growth, you will see that in the last 20 years migration rates have dropped significantly.
4. United states seems to always be in the mix for most migration across different decades. Interesting and relevant though, is Ukrain and Poland are leading this decade in absolute migration followed by USA, Pakistan, Saudi Arabia and Romania. Which all seems consistent with world events.