---
title: "Project 1"
output: html_document
author: 'Will Berritt'
date: "2024-02-17"
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## R Markdown

### Intro: Given a txt block of data the contains chess tournament results, the objective is to extract different pieces of information using pattern matching and create a new table (and CSV) that contains only specific and relevant pieces of information.

##### Import Packages
```{r}
# Packages
library(dplyr)
library(stringr)
```


##### Create DF with the txt file (uploaded to Github)
```{r read}
# Read txt file into data frame in R
txt_url <- "https://raw.githubusercontent.com/wberritt913/CUNY_DATA607/main/tournamentinfo.txt"
df1 <- read.table(url(txt_url), header = TRUE, sep = '\n')

# Check the dimensions and preview the data
dim(df1)
head(df1, 30)
```


##### Extract player names
```{r}
# Pattern match to extract the player names and create a data frame containing only their names
player_info <- data.frame(value=unlist(str_extract_all(toupper(df1), '[A-Z]+\\s[A-Z]+\\s[A-Z]*\\s?.?[A-Z]*')))
player_info <- player_info %>% slice(-1:-2)
player_info$value <- trimws(player_info$value)
head(player_info, 15)
```


##### Extract player states
```{r}
# Pattern match to extract the player states and create a data frame containing only their states
states_df <- data.frame(value=unlist(str_extract_all(toupper(df1), '[^A-Z]\\s[A-Z][A-Z]\\s')))
states_df$value <- trimws(states_df$value)
# Preview
head(states_df, 15)
```



##### Extract player total points
```{r}
# Pattern match to extract the player total points and create a data frame containing only their total points
total_points <- data.frame(value = unlist(points <- str_extract_all(toupper(df1), '[0-9]\\.[0-9]')))
total_points$value <- trimws(total_points$value)
head(total_points, 15)
```


##### Extract player ratings before + fix class
```{r}
# Pattern match to extract the player player ratings before and create a data frame containing only their player ratings before and check class
p_rating_before <- data.frame(value = unlist(str_extract_all(data.frame(value = unlist(str_extract_all(toupper(df1), '(:\\s\\s?[0-9])[0-9][0-9]?.?[0-9]?[0-9]?'))), '[0-9][0-9][0-9][0-9]?')))
class(p_rating_before$value)
# Change class
p_rating_before$value <- as.numeric(gsub(",", "", p_rating_before$value))
head(p_rating_before, 15)
```


##### Extract opponents' ratings
```{r}
# Pattern match to extract the player opponents' ratings and create a data frame containing only their opponents' ratings
opponents_df <- str_extract_all(str_extract_all(str_extract_all(str_extract_all(df1, '\\|...............................................\\|'), '\\|[0-9]..............................................\\|'), '[A-Z]....\\|'), '..\\|')
opponents_df <- data.frame(value = unlist(opponents_df))
opponents_df$value <- gsub("\\|", "", opponents_df$value)
colnames(opponents_df)[colnames(opponents_df) == "value"] <- "index"
opponents_df$index <- as.numeric(opponents_df$index)
head(opponents_df, 15)
```


##### Create main data frame
```{r}
# Combine all data frames into one data frame (keeping player_info as base data frame)
player_info$states <- states_df
player_info$total_points <- total_points
player_info$p_rating_before <- p_rating_before
player_info$index <- 1:nrow(player_info)
head(player_info, 15)
```



##### Join opponents' rating data to main data frame
```{r}
# Join the opponents data frame on the index
df3 <- left_join(opponents_df, player_info, by = 'index')
head(df3, 15)
```


##### Find average rating for each player's opponents
```{r}
# Get average of every 7 rows to calculate opponent average player rating before
groupings <- df3$p_rating_before %>%
  group_by(group = (row_number() - 1) %/% 7) 
averages <- groupings %>%
  summarise(avg_p_rating_before = round(mean(value, na.rm = TRUE)))
head(averages, 15)
```



##### Create new final data frame
```{r}
# Add the average to the main data frame and remove the index common key
player_info$average_opponent_rating <- averages$avg_p_rating_before
player_info <- player_info %>% select(-index)
player_data <- player_info
player_data
```


##### Check classes for export - fix problematic classes
```{r}
# Change the column classes
lapply(player_data, class)
# Change column classes + names to create final data frame
player_data$states <- as.character(player_data$states$value)
player_data$total_points <- as.numeric(player_data$total_points$value)
player_data$p_rating_before <- as.numeric(player_data$p_rating_before$value)
player_data <- player_data[, c("value", "states", "total_points", "p_rating_before", "average_opponent_rating")]
# Check class again
lapply(player_data, class)
# Rename value column
colnames(player_data)[colnames(player_data) == "value"] <- "name"
```

##### Create CSV
```{r}
# Create CSV
write.csv(player_data, "player_data_new1.csv", row.names = FALSE, quote = FALSE)
```

### Conclusion: Pattern matching was originally difficult as I was trying to use one pattern to extract all the relevant info. Later realizing it may make more sense to layer patten matching over each other to iteratively extract the data. The appending of the data frames was also tricky. It's important to verify the details of the data frames before combining them to ensure the final data frame is functional and contains the relevant information. This project is a strong testament the value of pattern matching and how it can be applied.

