---
title: "Project 2: NYC Transportation"
output: html_document
author: 'Will Berritt'
date: "2024-03-04"
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction: This data set includes NYC transportation usership data broken out by year, beginning in March 2020, just before the pandemic. I aimed to analyze a few different aspects of this data:
1. Which transportation methods had the greatest change in ridership from pre-pandemic to now?
2. May a potential change in population over time effect our perception of these numbers?
3. How do we compare now to pre-pandemic? Has our recovery to pre-pandemic rates plateaued?
With these questions in mind, I began working with and manipulating the data to extract these insights.

#### Load Packages
```{r, message=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(ggrepel)
```

#### Ridership: Read Data Set number 3
```{r, message=FALSE}
# Read MTA daily ridership data from CSV file
df3 <- read.csv('/Users/williamberritt/Downloads/MTA_Daily_Ridership_Data__Beginning_2020_20240228.csv')
```

#### Create separate pivoted data frames then join data back together
```{r, message=FALSE}
# Select relevant columns containing 'Estimated' and 'Total'
rider_ship <- df3 %>%
  select(Date, matches('Estimated|Total'))

# Reshape data from wide to long format using pivot_longer
rider_ship <- rider_ship |> pivot_longer(cols = (Subways..Total.Estimated.Ridership:Staten.Island.Railway..Total.Estimated.Ridership),
                           names_to = 'transport_method',
                           values_to ='value')

# Display the first few rows of the reshaped data frame
head(rider_ship)

# Assign transport methods based on pattern matching
rider_ship$transport_method <- ifelse(str_detect(rider_ship$transport_method, 'Subways'), 'subway', 
                                       ifelse(str_detect(rider_ship$transport_method, 'Buses'), 'buses', 
                                              ifelse(str_detect(rider_ship$transport_method, 'LIRR'), 'lirr', 
                                                     ifelse(str_detect(rider_ship$transport_method, 'Metro'), 'mnr', 
                                                            ifelse(str_detect(rider_ship$transport_method, 'Island'), 'staten',
                                                                   ifelse(str_detect(rider_ship$transport_method, 'Access'), 'access_a_ride',
                                                                          ifelse(str_detect(rider_ship$transport_method, 'Bridges'), 'bridges_tunnels', NA)))))))

# Select relevant columns for pre-pandemic data
pre_pandemic <- df3 %>%
  select(Date, contains("Pre"))

# Reshape pre-pandemic data from wide to long format using pivot_longer
pre_pandemic <- pre_pandemic |> pivot_longer(cols = (Subways....of.Comparable.Pre.Pandemic.Day:Staten.Island.Railway....of.Comparable.Pre.Pandemic.Day),
                                             names_to = 'transport_method',
                                             values_to ='value')

# Assign transport methods based on pattern matching
pre_pandemic$transport_method <- ifelse(str_detect(pre_pandemic$transport_method, 'Subways'), 'subway', 
                                       ifelse(str_detect(pre_pandemic$transport_method, 'Buses'), 'buses', 
                                              ifelse(str_detect(pre_pandemic$transport_method, 'LIRR'), 'lirr', 
                                                     ifelse(str_detect(pre_pandemic$transport_method, 'Metro'), 'mnr', 
                                                            ifelse(str_detect(pre_pandemic$transport_method, 'Island'), 'staten',
                                                                   ifelse(str_detect(pre_pandemic$transport_method, 'Access'), 'access_a_ride',
                                                                          ifelse(str_detect(pre_pandemic$transport_method, 'Bridges'), 'bridges_tunnels', NA)))))))

# Merge the two data frames on transport method and date
base_df <- left_join(rider_ship, pre_pandemic, by=c('transport_method', 'Date'))
base_df <- na.omit(base_df)

# Display the first few rows of the merged data frame
head(base_df)
```

#### Create new date variables and rename columns
```{r, message=FALSE}
# Rename columns
colnames(base_df) <- c('date', 'transport_method', 'ridership', 'percent_change')

# Convert 'date' to Date format
base_df$date <- as.Date(base_df$date, format = "%m/%d/%Y")

# Extract month, year, and day from 'date'
base_df$month <- format(base_df$date, "%m")
base_df$year <- format(base_df$date, "%Y")
base_df$day <- weekdays(base_df$date)
```

#### What do trends in ridership look like for each method?
```{r, message=FALSE}
# Plot ridership by data for each method
subway <- subset(base_df, transport_method=='subway')
plot(subway$date, subway$ridership)

buses <- subset(base_df, transport_method=='buses')
plot(buses$date, buses$ridership)

lirr <- subset(base_df, transport_method=='lirr')
plot(lirr$date, lirr$ridership)

mnr <- subset(base_df, transport_method=='mnr')
plot(mnr$date, mnr$ridership)

staten <- subset(base_df, transport_method=='staten')
plot(staten$date, staten$ridership)

access_a_ride <- subset(base_df, transport_method=='access_a_ride')
plot(access_a_ride$date, access_a_ride$ridership)

bridges_tunnels <- subset(base_df, transport_method=='bridges_tunnels')
plot(bridges_tunnels$date, bridges_tunnels$ridership)
```

#### Create data frame with for total ridership grouped by month
```{r, message=FALSE}
# Group data by year and month, calculating total ridership
total_ridership_by_month <- base_df |> group_by(year, month) |> 
  summarize(total_ridership = sum(ridership))
```

#### Compare first Wednesday to average of last 3 Wednesdays (only using first Wednesday because rates started to decline quickly in March. The first Wednesday seems like the best proxy of pre-pandemic behavior)
```{r, message=FALSE}
# Select data for the first Wednesday of March 2020
first_wednesday <- subset(base_df, day == 'Wednesday' & date == '2020-03-04')

# Select data for all Wednesdays
last_wednesday <- subset(base_df, day == 'Wednesday')
last_wednesday <- tail(last_wednesday, 21)

# Calculate average ridership for the last 3 Wednesdays by transport method
avg_last_3_wed <- last_wednesday |> group_by(transport_method) |> 
  summarize(avg_ridership_last_3 = mean(ridership))

# Merge data for the first Wednesday and average ridership for the last 3 Wednesdays
combined_wednesday <- left_join(first_wednesday, avg_last_3_wed, by = 'transport_method')

# Calculate total difference and percentage difference in ridership
combined_wednesday$total_difference <- combined_wednesday$ridership - combined_wednesday$avg_ridership_last_3
combined_wednesday$percent_diff_wed <- (combined_wednesday$avg_ridership_last_3 - combined_wednesday$ridership) / combined_wednesday$ridership

# Subset data for relevant columns
change_in_ridership <- subset(combined_wednesday, select = c('transport_method', 'total_difference', 'percent_diff_wed'))

# Display the results, ordered by total difference
change_in_ridership[order(-change_in_ridership$total_difference),]

# Display the results, ordered by percentage difference
change_in_ridership[order(change_in_ridership$percent_diff_wed),]
```

#### Adjust by considering recent NYC population decline
```{r, message=FALSE}
# Create a data frame for NYC population
nyc_pop <- data.frame(
  year = c(2020, 2021, 2022, 2023, 2024),
  population = c(8700000, 8500000, 8300000, 8200000, 8100000))

# Convert 'year' to numeric type
nyc_pop$year <- as.numeric(nyc_pop$year)

# Convert 'year' to numeric type in the base data frame
base_df$year <- as.numeric(base_df$year)

# Merge base data frame with NYC population data
ridership_population <- left_join(base_df, nyc_pop, by='year')

# Group data by year, calculating yearly ridership and daily population sum
ridership_population <- ridership_population |> group_by(year) |> 
  summarize(yearly_ridership = sum(ridership), daily_pop_sum = sum(population))

# Calculate the rate of use per 1000 people
ridership_population$rate_of_use <- (ridership_population$yearly_ridership / ridership_population$daily_pop_sum) * 1000

# Group data by year, calculating the mean ridership rate per 1000 people
ridership_by_year_by_transport <- ridership_population |> group_by(year) |> 
  summarize(ridership_rate_per_1000 = mean(rate_of_use))

# Convert 'year' to numeric type
ridership_by_year_by_transport$year <- as.numeric(ridership_by_year_by_transport$year)

# Plot the ridership rate per 1000 people over the years
plot(ridership_by_year_by_transport$year, ridership_by_year_by_transport$ridership_rate_per_1000)

```

#### Get ridership data grouped by year adjusted for population to see how different it is now vs pre pandemic
```{r, message=FALSE}
# Select data for the first week of March 2020
first_week_march_rate <- subset(base_df, date %in% c('2020-03-01', '2020-03-02', '2020-03-03', '2020-03-04', '2020-03-05', '2020-03-06', '2020-03-07'))

# Merge with NYC population data
first_week_march_rate <- left_join(first_week_march_rate, nyc_pop, by='year')

# Convert 'date' to numeric type
first_week_march_rate$date <- as.numeric(first_week_march_rate$year)

# Group data by year, calculating yearly ridership and daily population sum for the first week of March
march_grouping <- first_week_march_rate |> group_by(year) |> 
  summarize(ridership = sum(ridership), daily_pop_sum = sum(population))

# Calculate the ridership rate per 1000 people for the first week of March
march_grouping$ridership_rate_per_1000 <- (march_grouping$ridership / march_grouping$daily_pop_sum) * 1000

# Subset relevant columns
march_grouping <- subset(march_grouping, select=c('year', 'ridership_rate_per_1000'))

# Combine data for the first week of March with the yearly ridership rate per 1000 people
march_and_covid_ridership_pop_adjusted <- union_all(march_grouping, ridership_by_year_by_transport)

# Adjust the ridership rate for the entire year by multiplying by 7 (number of days in a week)
march_and_covid_ridership_pop_adjusted$ridership_rate_per_1000 <- march_and_covid_ridership_pop_adjusted$ridership_rate_per_1000 * 7

# Display the adjusted data
march_and_covid_ridership_pop_adjusted
```

# Conclusion:
1. The subway took the greatest hit in terms of gross ridership but the Staten Island Railway lost the greatest portion of riders.
2. Due to a nearly 7% decrease in NYC population, these numbers appear more significant than they really are. We can estimate that in theory, ridership should be 7% lesser today than pre-pandemic due to population decrease. However, even with population adjusted for, we are seeing a significantly greater drop than 7%.
3. We seem to have plateaued in our recovery to pre-pandemic levels of ridership. Achieving a ridership rate of 61.8% in 2022, 68.4% in 2023 and now back down to 64.8% in 2024. The relatively small amounts of change in recent times leads me to believe that we are now in a steady state. It's also possible that 2024 data is skewed by other factors like season or weather or other circumstances.