---
title: "Week 7 Assignment"
output: html_document
date: "2024-03-10"
author: 'Will Berritt'
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Libraries
library(xml2)
library(stringr)
library(tidyverse)
library(rvest)
library(jsonlite)
library(purrr)
```

# Intro: The goal was to create HTML, XML, and JSON files that I could then read from in R Studio. Using correct syntax in the different files is key to easily reading the relevant book data that I manually entered.

### Extract HTML info
```{r }
# Read HTML file
html <- read_html('https://github.com/wberritt913/CUNY_DATA607/blob/main/book_info.html')

# Extract text content from specific HTML elements
html_snip <- html |> html_elements("body") |> html_elements("p") |> html_elements("span") |> html_text2()

# Extract information using regular expressions
headers <- str_extract_all(html_snip, "<td>(.*?)</td>")
filtered_headers <- headers[lengths(headers) > 0]

# Extract titles, authors, and attributes using sequence and replacement
title <- filtered_headers[seq(1, length(filtered_headers), by = 3)]
title <- str_replace_all(title, "<td>(.*?)</td>", "\\1")

authors <- filtered_headers[seq(2, length(filtered_headers), by = 3)]
authors <- str_replace_all(authors, "<td>(.*?)</td>", "\\1")

attributes <- filtered_headers[seq(3, length(filtered_headers), by = 3)]
attributes <- str_replace_all(attributes, "<td>(.*?)</td>", "\\1")

# Create a tibble (data frame) with the extracted information
books_html <- tibble(
  title = title,
  authors = authors,
  attributes = attributes
)

# Display the tibble
books_html

```

### Extract XML info
```{r, echo=FALSE}
# Read XML file
xml_file <- read_xml('https://raw.githubusercontent.com/wberritt913/CUNY_DATA607/main/week7xml.xml')

# Extract text content from nested XML elements
book <- xml_file |> xml_find_all("w:body") |> xml_find_all("wx:sect") |> xml_find_all("w:p") |> xml_find_all("w:r") |> xml_find_all("w:t") |> xml_text()

# Use regular expressions to extract information
code_snip <- str_view_all(book, "<books>(.*?)</books>")
title <- str_extract_all(code_snip, "<title>(.*?)</title>")
filtered_titles <- title[lengths(title) > 0]
title <- str_replace_all(filtered_titles, "<title>(.*?)</title>", "\\1")

authors <- str_extract_all(code_snip, "<authors>(.*?)</authors>")
filtered_authors <- authors[lengths(authors) > 0]
authors <- str_replace_all(filtered_authors, "<authors>(.*?)</authors>", "\\1")

attributes <- str_extract_all(code_snip, "<attributes>(.*?)</attributes>")
filtered_attributes <- attributes[lengths(attributes) > 0]
attributes <- str_replace_all(filtered_attributes, "<attributes>(.*?)</attributes>", "\\1")

# Create a tibble (data frame) with the extracted information
books_xml <- tibble(
  title = title,
  authors = authors,
  attributes = attributes
)

# Display the tibble
books_xml

```
### Extract JSON info
```{r}
# Read JSON file
json_data <- fromJSON("https://raw.githubusercontent.com/wberritt913/CUNY_DATA607/main/week7_json.json")

# Display the content of the JSON data
json_data

```
### I had trouble storing the HTML and XML codes in the correct files. I took the code and pasted them into different IDEs and it would return the intended tabular format but when saving the files and opening them, it would just open in the web browser as the actual code written. Here, I created 'files' to use the formats and commands as initially intended.
```{r}
# Write HTML content to file
writeLines('
<html>
<head>
    <title>Book Title</title>
</head>
<body>
    <table>
        <tr>
            <th>title</th>
            <th>authors</th>
            <th>attributes</th>
        </tr>
        <tr>
            <td>The Lightning Thief</td>
            <td>Rick Riordan</td>
            <td>Fantasy, Greek mythology, Adventure</td>
        </tr>
        <tr>
            <td>Old Yeller</td>
            <td>Fred Gipson</td>
            <td>Classic, Coming-of-age, Adventure</td>
        </tr>
        <tr>
            <td>The Da Vinci Code</td>
            <td>Dan Brown, Jeff Harding, Paul Michael</td>
            <td>Mystery, Thriller, Conspiracy</td>
        </tr>
    </table>
</body>
</html>', "books.html")  # Write the HTML content to a file named "books.html"


# Write XML content to file
writeLines('
<books>
  <book>
    <title>The Lightning Thief</title>
    <authors>Rick Riordan</authors>
    <attributes>Fantasy, Greek mythology, Adventure</attributes>
  </book>
  <book>
    <title>Old Yeller</title>
    <authors>Fred Gipson</authors>
    <attributes>Classic, Coming-of-age, Adventure</attributes>
  </book>
  <book>
    <title>The Da Vinci Code</title>
    <authors>Dan Brown, Jeff Harding, Paul Michael</authors>
    <attributes>Mystery, Thriller, Conspiracy</attributes>
  </book>
</books>', "books.xml")  # Write the XML content to a file named "books.xml"


# Write JSON content to file
writeLines('{
  "books": [
    {
      "title": "The Lightning Thief",
      "authors": ["Rick Riordan"],
      "attributes": ["Fantasy", "Greek mythology", "Adventure"]
    },
    {
      "title": "Old Yeller",
      "authors": ["Fred Gipson"],
      "attributes": ["Classic", "Coming-of-age", "Adventure"]
    },
    {
      "title": "The Da Vinci Code",
      "authors": ["Dan Brown", "Jeff Harding", "Paul Michael"],
      "attributes": ["Mystery", "Thriller", "Conspiracy"]
    }
  ]
}', "books.json")  # Write the JSON content to a file named "books.json"

```
### Extract HTML info pt. II
```{r}
# Read HTML file
html <- read_html('books.html')

# Extract 'table' element and convert to a data frame
table <- html |> html_element("table") |> html_table()

# Display the resulting data frame
table
```
### Extract XML info pt. II
```{r}
# Read XML file
xml <- read_xml('books.xml')

# Extract 'book' elements from XML
book <- xml |> xml_find_all("book")

# Create a data frame (tibble) with extracted information
xml_df <- tibble(
  title = book |> xml_find_all("title") |> xml_text(),
  authors = book |> xml_find_all("authors") |> xml_text(),
  attributes = book |> xml_find_all("attributes") |> xml_text()
)

# Display the resulting data frame
xml_df
```
### Extract JSON info pt. II
```{r}
# Read JSON file
json_data <- fromJSON("books.json")

# Create a data frame (tibble) with extracted information
json_df <- tibble(
  titles = json_data$books$title,
  authors = sapply(json_data$books$authors, paste, collapse = ", "),
  attributes = sapply(json_data$books$attributes, paste, collapse = ", ")
)

# Display the original JSON data
json_data

# Display the resulting data frame
json_df

```
# Conclusion: The JSON data appears to be easily read in both cases and returns a table without much manipulation. The table appears different from the HTML and XML files but those also required more code to extract and formulate a table.

